Bias correction in GLMs

### 2021-07-15 bias_correction.rmd
Normal for age
Population averaging seems to work

We perform a simple simulation for a fixed effect model
\begin{align*}
\mathrm{logit(status} = 1) &= \eta \\ 
\eta &= \beta_0 + \beta_{\mathrm{A}} \mathrm{Age} + \beta_{\mathrm{W}} \mathrm{Wealthindex} \\
\mathrm{Age} &\sim \mathrm{Uniform}(0.2, 1) \\
\mathrm{Wealthindex} &\sim \mathrm{Normal}(0, 1) \\
\beta_0 &= 0.7 \\
\beta_{\mathrm{A}} &= 0.3 \\
\beta_{\mathrm{W}} &= 0.6
\end{align*}

@@

age_max <- 1
age_min <- 0.2
age <- runif(N, age_min, age_max)

@@

Variable effect plots -- varpred and population averaging approach

```{r simple_vareffects} 
# Age
## varpred way
simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="varpred")

# Wealth index
## varpred - no bias adjustment
simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="varpred")

## Bias adjusted
simple_vareff_wealthindex_adjust <- varpred(simple_mod, "wealthindex", isolate=TRUE, pop.ave=TRUE, modelname="Pop. ave")

vareff_wealthindex <- simple_vareff_wealthindex
vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_adjust$preds))
wealthindex_plot <- (plot(vareff_wealthindex)
 	+ labs(y="", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ scale_colour_manual(values=c("black", "blue"))
	+ theme(legend.position="bottom")
)
```

@@

```{r pop_ave_fun}
popavefun <- function(mod, focal, non.focal, level=0.95, modelname="Pop. ave", ...) {
	mf <- model.matrix(mod)
	mm <- (mf
		%>% data.frame()
		%>% mutate_at(non.focal, mean)
		%>% as.matrix()
	)
	vc <- vcov(mod)
	linpred <- as.vector(mm %*% coef(mod))
	pse_var <- sqrt(rowSums(mm * t(tcrossprod(data.matrix(vc), mm))))
	z.val <- qnorm(1 - (1 - level)/2)
	pred_df <- (mf
		%>% data.frame()
		%>% select_at(focal)
		%>% mutate(fit = linpred
			, lwr = plogis(fit - z.val*pse_var)
			, upr = plogis(fit + z.val*pse_var)
			, fit = plogis(fit)
			, model = modelname
			, se = NA
		)
	)
	return(pred_df)
}

@@

## Pop. average
simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave")

## Bias adjust
simple_vareff_age_adjust <- varpred(simple_mod, "age", isolate=TRUE, bias.adjust=TRUE, modelname = "Bias adjust")

vareff_age <- simple_vareff_age
vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_pop, simple_vareff_age_adjust$preds))
age_plot <- (plot(vareff_age)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ scale_colour_manual(values=c("black", "blue", "red"))
	+ theme(legend.position="bottom")
)
```

@@

```{r simple_vareff_wealthindex, fig.cap = "A comparison of population averaged and varpred-based predictions. We also apply bias-adjustemnt to the `vapred` predictions. For `age`, we implement the naive approach to compute the predictions in `popavefun` function and then implement the same in `vareffects` so as to use the centering machineries. In both cases, the population averaging and varpred gives similar estimates. The estimated population average is very close to the observed in the case of `wealthindex` but slightly higher in the case of `age` for the unadjusted estimates (but very low for bias-adjusted). , see previous paragraph."}
# Wealth index
## varpred 
simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="varpred")

## Pop. average
simple_vareff_wealthindex_pop <- varpred(simple_mod, "wealthindex", isolate=TRUE, pop.ave=TRUE, modelname="Pop. ave")

## Bias adjust
simple_vareff_wealthindex_adjust <- varpred(simple_mod, "wealthindex", isolate=TRUE, bias.adjust=TRUE, modelname="Bias adjust")

vareff_wealthindex <- simple_vareff_wealthindex
vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_pop$preds, simple_vareff_wealthindex_adjust$preds))
wealthindex_plot <- (plot(vareff_wealthindex)
 	+ labs(y="", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ scale_colour_manual(values=c("black", "blue", "red"))
	+ theme(legend.position="bottom")
)

@@

- age: unadjusted `r mean(simple_vareff_age_pop$fit)`; adjusted `r mean(simple_vareff_age_adjust$preds$fit)`
- wealthindex: unadjusted `r mean(simple_vareff_wealthindex_pop$preds$fit)`; adjusted `r mean(simple_vareff_wealthindex_adjust$preds$fit)`

>>>>

We perform a simple simulation for a fixed effect model
\begin{align*}
\mathrm{logit(status} = 1) &= \eta \\ 
\eta &= \beta_0 + \beta_{\mathrm{A}} \mathrm{Age} + \beta_{\mathrm{W}} \mathrm{Wealthindex} \\
\mathrm{Age} &\sim \mathrm{Normal}(0.2, 1) \\
\mathrm{Wealthindex} &\sim \mathrm{Normal}(0, 1) \\
\beta_0 &= 0.7 \\
\beta_{\mathrm{A}} &= 0.3 \\
\beta_{\mathrm{W}} &= 0.6
\end{align*}

@@

age_max <- 1
age_min <- 0.2
# age <- runif(N, age_min, age_max)
age <- rnorm(N, age_max, age_max)

@@

@@

```{r pop_ave_fun}
popavefun <- function(mod, focal, non.focal, level=0.95, steps=100, bins=50, modelname="Pop. ave", ...) {
	mf <- model.frame(mod)
	mm <- (mf
		%>% select_at(c(focal, non.focal))
	)
	check_df <- (mf
		%>% arrange_at(focal)
		%>% mutate(bin=ceiling(row_number()*bins/nrow(.)))
		%>% group_by(bin)
		%>% summarise_all(mean)
		%>% mutate(model="binned")
	)
	quant <- seq(0, 1, length.out=steps)
	mm <- sapply(c(focal, non.focal), function(x)quantile(mm[,x], quant), simplify = FALSE)
	mm <- do.call("expand.grid", mm)
	vc <- vcov(mod)
	mm2 <- model.matrix(formula(mod)[c(1,3)], mm)
	linpred <- as.vector(mm2 %*% coef(mod))
	pse_var <- sqrt(rowSums(mm2 * t(tcrossprod(data.matrix(vc), mm2))))
	z.val <- qnorm(1 - (1 - level)/2)
	pred_df <- (mm
		%>% select_at(focal)
		%>% mutate(fit = linpred
			, lwr = plogis(fit - z.val*pse_var)
			, upr = plogis(fit + z.val*pse_var)
			, fit = plogis(fit)
		)
		%>% group_by_at(focal)
		%>% summarise_at(c("fit", "lwr", "upr"), mean)
		%>% mutate(model = modelname, se=NA)
	)
	return(list(preds=pred_df, check=check_df))
}

@@

## Pop. average
simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave") 
binned_df <- simple_vareff_age_pop$check

vareff_age <- simple_vareff_age
vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_pop$preds))
age_plot <- (plot(vareff_age)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
	+ scale_colour_manual(values=c("black", "blue", "red"))
	+ theme(legend.position="bottom")
)
```

@@

```{r simple_vareff_wealthindex, fig.cap = "A comparison of population averaged and varpred-based predictions. We also apply bias-adjustemnt to the `vapred` predictions. For `age`, we implement the naive approach to compute the predictions in `popavefun` function and then implement the same in `vareffects` so as to use the centering machineries. In both cases, the population averaging and varpred gives similar estimates. The estimated population average is very close to the observed in the case of `wealthindex` but slightly higher in the case of `age` for the unadjusted estimates (but very low for bias-adjusted). , see previous paragraph."}
# Wealth index
## varpred 
simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=FALSE, modelname="varpred")

## Pop. average
simple_vareff_wealthindex_pop <- popavefun(simple_mod, "wealthindex", "age", modelname="Pop. ave")
binned_df <- simple_vareff_wealthindex_pop$check

vareff_wealthindex <- simple_vareff_wealthindex
vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_pop$preds))
wealthindex_plot <- (plot(vareff_wealthindex)
 	+ labs(y="", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
	+ scale_colour_manual(values=c("black", "blue", "red"))
	+ theme(legend.position="bottom")
)

@@

- age: pop. average `r mean(simple_vareff_age_pop$preds$fit)`; varpred `r mean(simple_vareff_age$preds$fit)`
- wealthindex: pop. average `r mean(simple_vareff_wealthindex_pop$preds$fit)`; varpred `r mean(simple_vareff_wealthindex$preds$fit)`
